---
title: "Demos"
author: "Cory Merow"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{An overview of making rangeModelMetadata objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

```{r}
library(maskRanger)
library(raster)
library(virtualspecies)
```

# Example 1: Competing species

Make a demo raster of occurrence probabilities describing the potential distribution

```{r}
worldclim=getData("worldclim", var = "bio", res = 10)
# let's focus on South America
worldclim=crop(worldclim,c(-100,-20,-60,15))
```

```{r,fig=T}
spA.parameters=formatFunctions(bio1 = c(fun = 'dnorm', mean = 240, sd = 30),
                               bio12 = c(fun = 'dnorm', mean = 3000, sd = 1000))
spA.tmp=generateSpFromFun(raster.stack = worldclim[[c("bio1", "bio12")]],
                          parameters = spA.parameters)
spA=spA.tmp$suitab.raster
plot(spA)
```

Make a demo mask. let's pretend this species, spB, is a competitor of the first species, spA, and that it excludes spA anywhere it occurs

```{r}
spB.parameters=formatFunctions(bio1 = c(fun = 'dnorm', mean = 250, sd = 20),
                               bio12 = c(fun = 'dnorm', mean = 4000, sd = 1000))
spB.tmp=generateSpFromFun(raster.stack = worldclim[[c("bio1", "bio12")]],
                          parameters = spB.parameters)
spB=spB.tmp$suitab.raster
# let's focus on South America
plot(spB)

```

```{r,fig.width=7,fig.height=4}
spAnoB=maskRanger(potentialDist=spA,
                    maskLayers=spB,
                    logicString='maskLayers>0.5',
                    method='mask')
plot(stack(spA,spB,spAnoB))
```


It also works with multiple time steps. You can either have a different mask for each time step as we show below (e.g., you might have a potential distribution for each decade as well as a forest cover mask for each decade) or a single mask to apply to each decade (e.g., elevation, which probably shouldn't be changing each decade.)

```{r,fig.width=7,fig.height=7}
# make a demo raster of occurrence probabilities describing the potential distribution over 9 time steps
spA.t=stack(spA,spA-.1,spA+.1)
spB.t=stack(spB,spB+.1,spB-.1)
logicString='maskLayers>0.5'
spAnoB.t=maskRanger(potentialDist=spA.t,maskLayers=spB.t,logicString,method='mask')
plot(stack(spA.t,spB.t,spAnoB.t))
```

You can also specify any sort of arbitrary conditional statement (although this might note make sense for competing species)

```{r,fig.width=7,fig.height=7}
# make a demo raster of occurrence probabilities describing the potential distribution over 9 time steps
logicString='maskLayers>0.5 & maskLayers<0.7'
spAnoB.t2=maskRanger(potentialDist=spA.t,maskLayers=spB.t,logicString,method='mask')
plot(stack(spA.t,spB.t,spAnoB.t2))
```


<!-- ```{r} -->
<!-- spA.bin=spA.tmp$suitab.raster>0.5 -->

<!-- spB.bin=spB.tmp$suitab.raster>0.5 -->
<!-- ``` -->

